Secure Composition of Robust and Optimising Compilers

### 2025-02-13 icfp/main.tex
comments
move

% Fixing: Too many math alphabets used in version normal
\newcommand\hmmax{0}
\newcommand\bmmax{0}
\documentclass[acmsmall]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

\input{sections/intro.tex}

\input{sections/background.tex}

\section{Secure Composition}\label{sec:sequential}

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g.,~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls*{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls*{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls*{cf} and \gls*{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls*{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls*{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls*{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls*{cf} and \gls*{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls*{cf} or \gls*{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.

\smallskip

To analyse the security of compilation passes and their interaction within a compilation pipeline, we rely on a few key notions: the definition of a trace relation, and the definition of when is a trace relation well-formed with respect to a class.
%
Consider traces $\src{\trace}$ and $\trg{\trace}$ as well as two trace relations $\sim_1$ and $\sim_2$. 
% 
The traces are related $\src{\trace}\sim_1\bullet\sim_2\trg{\trace}$ if there is another trace $\irl{\trace}$ such that $\src{\trace}\sim_1\irl{\trace}$ and $\irl{\trace}\sim_2\trg{\trace}$.
% 
\bul{A relation $\sim$ is well-formed w.r.t. a class of target-level properties $\trg{\class}$} iff \rul{the universal image preserves set membership}.
\begin{definition}[Well-formedness of $\sim$ for a Class $\trg{\class}$]\label{def:wfc:sig:tracerel}
  % $$

  \begin{nscenter}
  \noindent
  \text{\bul{$\wfcsig{\sim}{\trg{\class}}$}} \isdef \text{\rul{$\forall \trg{\pi}\in\trg{\class}, \sigma_\sim(\trg{\pi})\in\sigma_\sim(\trg{\class})$}}
  \end{nscenter}
  % $$
\end{definition}

We can now state our main result: secure compilers in the robust compilation framework~\cite{abate2021extacc} compose {\em sequentially}. 
%

Let \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserve the class $\sigma_{\sim_2}\left(\obj{\class[_1]}\right)$ under $\sim_1$} and let \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserve the class $\obj{\class[_2]}$ under $\sim_2$}.
Then, \iul{when the cross-language relation $\sim_2$ is well-formed w.r.t. class $\obj{\class[_1]}$}, it follows that \oul{the composed compiler\footnote{Here, $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ means first running $\cc{\src{L}}{\trg{L}}$ and then $\cc{\trg{L}}{\obj{L}}$.} $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\obj{\class[_1]}\cap\obj{\class[_2]}$ under $\sim_1\bullet\sim_2$}.
% 
%For space reasons, \Thmref{thm:urtp} and \Thmref{thm:lrtp} are not presented here, but can be found in the technical appendix ($\CoqSymbol$).
% Note that it is not necessary to require $\sim_2$ to be well-formed w.r.t. $\src{\class[_1]}$, since the 
% \MPin{why}
\begin{theorem}[Composition of Secure Compilers w.r.t. $\sigma$]\label{thm:rtpsim:sig}
  $\;$

  If \bul{$\rtpsig{\cc{\src{L}}{\trg{L}}}{\tilde{\sigma}_{\sim_2}\left(\obj{\class[_{1}]}\right)}{\sim_1}$}, \rul{$\rtpsig{\cc{\trg{L}}{\obj{L}}}{\obj{\class[_2]}}{\sim_2}$}, and \iul{$\wfcsig{\sim_2}{\obj{\class[_1]}}$}, \\ then \oul{$\rtpsig{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\obj{\class[_{1}]}\cap\obj{\class[_{2}]}}{\sim_1\bullet\sim_2}$}. \Coqed
\end{theorem}

Since the composition of secure compilers is again a secure compiler, the theorems generalise to a whole chain of $n$ secure compilers. 
% 
\Cref{thm:rtpsim:sig} can also be stated for the existential image $\tau_\sim\left(\src{\pi}\right)$, but in the interest of space that result has been moved to the appendix.
 % (\Cref{subsec:extimg}).
% 
Crucially, \Cref{thm:rtpsim:sig} also holds for \emph{classes of hyperproperties}, and thus, compilers that robustly preserve hyperproperties can be composed with each other as well as with compilers that robustly preserve properties.
% 
%Unfortunately, proofs of compilers that robustly preserve hyperproperties (including hypersafety) are scarce in the literature due to their complexity.
%\MP{ so what ? }
%  we don't discuss this further?
%

% It is furthermore possible to conflate the notion of classes and hyperproperties by intersecting all elements of a class of hyperproperties, to obtain one combined hyperproperty.
% This idea has been discussed in other work~\cite{clarkson2008hyper}.
%  MP: i think i stated the point of these 2 lines above

If we take SecurePtrs and FaCT from \Cref{subsec:bg:rtp} and compose them according to \Cref{thm:rtpsim:sig}, we obtain a compiler that robustly preserves the intersection of safety properties and the \gls*{cct} hyperproperty. 
That is, for a source component that robustly satisfies any set of safety properties {\em and} \gls*{cct}, the compiled target component also robustly satisfies the same set of safety properties and \gls*{cct}.
% The intersection simply uses the lifted version of the respective safety properties.
%\MP{meaning?}

Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid w.r.t. security, as long as the trace-relations have no effect on the respective classes.

So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\trg{\class[_{1}]}$ or $\trg{\class[_{2}]}$, respectively}, \rul{their corresponding well-formed trace-relations}, \iul{and indifference of the classes with respect to these trace relations}, \oul{for any order of their composition, the composed compiler robustly preserves the intersection of $\trg{\class[_{1}]}$ and $\trg{\class[_{2}]}$}.
% 
% \MPin{
% 	later on, say that optimisation passes remove elements from traces, so they should be fine w.r.t. the result with $\tau$.
% 	also, yes, if $\tau$ adds we have a problem, if it removes we don't
% \MKin{ Optimisations that remove elements from traces are not necessarily sound,
%  e.g., remove all deallocations to "optimise" memory management by offloading
%  the work to the OS.
% }
% }

\begin{corollary}[Swapping Secure Compiler Passes]\label{corr:swappable}
  $\;$

  If \bul{$\rtpsig{\cc[_{1}]{\trg{L}}{\trg{L}}}{\trg{\class[_{1}]}}{\sim_1}$ and $\rtpsig{\cc[_{2}]{\trg{L}}{\trg{L}}}{\trg{\class[_{2}]}}{\sim_2}$}, %
  \rul{$\wfcsig{\sim_1}{\trg{\class[_2]}}$ and $\wfcsig{\sim_2}{\trg{\class[_1]}}$}, %
  and \iul{$\tilde{\sigma}_{\sim_2}\left(\trg{\class[_1]}\right)=\trg{\class[_1]}$ as well as $\tilde{\sigma}_{\sim_1}\left(\trg{\class[_2]}\right)=\trg{\class[_2]}$},
  then \oul{$\rtpsig{\cc[_{1}]{\trg{L}}{\trg{L}}\circ\cc[_{2}]{\trg{L}}{\trg{L}}}{\trg{\class[_{1}]}\cap\trg{\class[_{2}]}}{\sim_1\circ\sim_2}$ and $\rtpsig{\cc[_{2}]{\trg{L}}{\trg{L}}\circ\cc[_{1}]{\trg{L}}{\trg{L}}}{\trg{\class[_{2}]}\cap\trg{\class[_{1}]}}{\sim_2\circ\sim_1}$}. \Coqed
\end{corollary}

Coming back to the example composing SecurePtrs with FaCT, it is likely the case that \Cref{corr:swappable} is not applicable.
While first running SecurePtrs and then FaCT should be fine, the other direction has potential security hazards, since the SecurePtrs compiler does not account for cryptographic-constant time primitives, such as \texttt{ctselect}.

\subsection{Secure Compiler Composition with Same Trace Models}
When the cross-language trace relation is an equality, \Cref{thm:rtpsim:sig} collapses:
Given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

\begin{corollary}[Composition of Secure Compilers]\label{corr:rtp}
  $\;$

  If \bul{$\rtp{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}$} and \rul{$\rtp{\cc{\trg{L}}{\obj{L}}}{\class[_{2}]}$}, then \oul{$\rtp{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}$}. \Coqed
\end{corollary}

\Cref{corr:rtp} provides an easy way to compose secure compilers without well-formedness of trace relations. 
% 
However, while \Cref{thm:rtpsim:sig} explicitly requires that $\sim_1$ is well-formed w.r.t. $\trg{\class[_2]}$, if $\sim_2$ is not well-formed w.r.t. $\trg{\class[_1]}$, care must be taken. 
%Worst-case is that the resulting class $\sigma_{\sim_1\bullet\sim_2}\left(\trg{\class[_1]}\cap\trg{\class[_2]}\right)$ becomes empty, even though $\trg{\class[_1]}\cap\trg{\class[_2]}$ is not necessarily empty.
This is further discussed in \Cref{subsec:compatsecpasses}.

%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
We can also obtain a specialised version of \Cref{corr:swappable}:
%So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition, the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.
% 
% \MPin{
% 	later on, say that optimisation passes remove elements from traces, so they should be fine w.r.t. the result with $\tau$.
% 	also, yes, if $\tau$ adds we have a problem, if it removes we don't
% \MKin{ Optimisations that remove elements from traces are not necessarily sound,
%  e.g., remove all deallocations to "optimise" memory management by offloading
%  the work to the OS.
% }
% }

\begin{corollary}[Swapping Secure Compiler Passes]\label{corr:swappable:one}
  $\;$

  If {$\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{1}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{2}]}$}, then {$\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}\circ\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{1}]\cap\class[_{2}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}\circ\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{2}]\cap\class[_{1}]}$}. \Coqed
\end{corollary}

\section{Security Properties Formalisation \& Composition\pages{1.5}}\label{sec:compprop}

This section introduces trace properties of interest for this paper: \gls*{tms}, \gls*{sms}, \gls*{ms}, \gls*{scct}, and \gls*{ss}.
These properties are of practical importance (as mentioned in \Cref{sec:introduction}) and also of interest in the case study presented later (\Cref{sec:casestud:defs,sec:casestud:rtp}). 
This section presents all of them, despite the fact that they are inspired by existing work, in order to showcase all that is required for a formal proof of security for a realistic compilation toolchain.
%\MP{ say something about this in the rebuttal: only the defs are borrowed, the universal images and all that work is new}
%\MK{ universal images are not new. }
% , none of the properties is novel, for self-containment of the paper, their definition is necessary for the case study.
% For each of the presented properties, the technical report defines corresponding monitors, which refine a given property and have a key role in the proofs of this paper. 
The technical report defines monitors for each of the presented properties.
% 
Monitors refine each property and have a key role in the proofs of this paper.
% managed to prove 
%  if tms-monitor --[trace]-->* ∅  
%  then trace ∈ tms
% but not
%  if trace ∈ tms
%  then tms-monitor --[trace]-->* ∅

\subsection{A Trace Model for Memory Safety}\label{subsec:basic:memsafety:tracemodel}

For simple memory safety composed of temporal and spatial memory safety, the trace model defines events ($\event[_{\mssafe}]$) as either the empty event ($\emptyevent$), a crash ($\lightning$), or a base-event ($\preevent[_{\mssafe}]$).

\vspace{-1.0em}
\[
  \begin{array}{rcll}
    \text{(Base-Events)}&\preevent[_{\mssafe}] &:=& \ev{Alloc \loc\ n} \mid \ev{Dealloc \loc} \mid \ev{Use \loc\ n} \\
    \text{(Events)}&\event[_{\mssafe}] &:=& \preevent[_{\mssafe}] \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]

Base-events describe the actual kind of event that happened.
For the basic memory-safety properties, these are three variants:
First, the allocation event ($\ev{Alloc\ \loc\ n}$) that fires whenever a program claims $n$ cells of memory and stores them at address $\loc$, where addresses are assumed to be unique.
Second, deallocation ($\ev{Dealloc\ \loc}$) announces that the object at location $\loc$ is freed.
Third, an event to describe reads from and writes to the $n$-th memory cell from address $\loc$ ($\ev{Use\ \loc\ n}$).

\subsubsection{Temporal Memory Safety}

\gls*{tms}~\cite{nagarakatte2010cets} is a safety property that describes that an unallocated object must not be (re-)used.

\begin{definition}[\glsfirst*{tms}]\label{def:trace:tmsdef}
  % \footnotesize
  $$
  \tmssafe:=\left\{\trace[_{\mssafe}] \left| \begin{array}{rcl}
    \ev{Alloc\ \loc\ n}&\le_{\trace[_{\mssafe}]}&\ev{Dealloc\ \loc} \\
    \ev{Use\ \loc\ n}&\le_{\trace[_{\mssafe}]}&\ev{Dealloc\ \loc} \\
    \text{at most one }\ev{Dealloc\ \loc}&\text{in}&\trace[_{\mssafe}] \\
    \text{at most one }\ev{Alloc\ \loc\ n}&\text{in}&\trace[_{\mssafe}] \\
  \end{array}\right.\right\}
  $$
\end{definition}
Hereby, the notation $\event[_{1}]\le_{\trace}\event[_{2}]$ means that if $\event[_{1}]$ is in $\trace$ and if $\event[_{2}]$ is in $\trace$, then $\event[_{1}]$ appears before $\event[_{2}]$.

\subsubsection{Spatial Memory Safety}

\gls*{sms}~\cite{nagarakatte2009soft} is a safety property that prohibits out-of-bounds accesses.

\begin{definition}[\glsfirst*{sms}]\label{def:trace:smsdef}
% \small
  % \begin{nscenter}

  \noindent
  \[
  \smssafe:=\left\{\trace[_{\mssafe}] \left|\begin{array}{rcl}
      \text{If }\ev{Alloc\ \loc\ n}\le_{\trace_{\mssafe}}\ev{Use\ \loc\ m}, \text{ then }m<n
  \end{array}\right.\right\}
  \]
  % \end{nscenter}
\end{definition}

\subsubsection{Memory Safety}

In spirit of earlier work~\cite{nagarakatte2009soft,nagarakatte2010cets,jim2002cyclone,necula2005ccured,michael2023mswasm}, full \gls*{ms} is the intersection of \Cref{def:trace:tmsdef,def:trace:smsdef}.

\begin{definition}[\glsfirst*{ms}]\label{def:trace:msdef}
  $
  \mssafe:=\tmssafe \cap \smssafe
  $
\end{definition}

Note that \Cref{def:trace:msdef} ignores data isolation, so there may still be memory-safety issues introduced by side-channels.

\subsection{A Trace Model for Memory Safety with Constant Time}\label{subsec:scct:tracemodel}

To express Constant Time, we extend the memory safety trace model with a {\em security tag} ($\securitytag{}$) that indicates whether events contain sensitive information ($\lock$) or not ($\unlock$).

\vspace{-.5em}
\[
  \begin{array}{rrcl}
    (\text{Base-Events}) & \preevent[_{\ctsafe}] &:=& \preevent[_{\mssafe}] \mid \ev{Branch\ }n \mid \ev{Binop\ }n\\
    (\text{Security Tags}) & \securitytag{} &:=& \lock \mid \unlock\\ 
    (\text{Events}) & \event[_{\ctsafe}] &:=& \preevent[_{\ctsafe}];\securitytag{} \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]

For cryptographic code, there is a general guideline that secrets must not be visible on a trace~\cite{ctguidelines}, i.e., secrets should not be marked as $\unlock$.
In turn, an instruction whose timing is data-dependent must not have a secret as an operand.
Typical operations with data-dependent timing are branches and certain binary operations, such as division.%
\footnote{
	This is architecture-dependent, but division is an operation that serves as a classic example for a data-dependent timing instruction~\cite[p.~755]{arm-refman}.
}
Both operations are represented in the trace model by extending the set of base-events with branches ($\ev{Branch\ n}$) and binary operations ($\ev{Binop\ n}$).

\subsubsection{Strict Cryptographic Constant Time}

\gls*{cct} is a hypersafety property~\cite{barthe2018sec} and, thus, difficult to check with monitors.
This is because, intuitively, hypersafety properties can relate multiple execution traces with each other, but monitors work on a single execution.
It is a common trick to sidestep this issue by means of overapproximation: this section defines the property \gls*{scct}, a stricter variant of \gls*{cct} (inspired by earlier work~\cite{almeida2017jasmin}) that enforces the policy that no secret appears on a trace.
Programs that satisfy \gls*{scct} also satisfy \gls*{cct}, but programs that satisfy \gls*{cct} may not satisfy \gls*{scct}.

\begin{definition}[\glsfirst*{scct}]\label{def:trace:scctdef}
% \begin{nscenter}

  \noindent\[
  \scctsafe:=\left\{\trace[_{\ctsafe}] \left|\begin{array}{l}
      \trace[_{\ctsafe}]=\hole{\cdot} \text{ or } \\\exists\trace[_{\ctsafe}'],\trace[_{\ctsafe}]=\preevent[_{\ctsafe}];\unlock\cdot\trace[_{\ctsafe}'] \wedge \trace[_{\ctsafe}']\in\scctsafe
    \end{array}\right.\right\}
  \]
% \end{nscenter}
\end{definition}

\gls*{scct} may appear overly strict, since it seems that secrets must not occur on a trace (since $\securitytag{}$ is forced to be $\unlock$). 
However, this is considered standard practice in terms of coding guidelines~\cite{ctguidelines}.
Moreover, programs that have been compiled with FaCT~\cite{cauligi2019fact} and run with a ``data independent timing mode''~\cite{arm-refman,intel-refman} enabled do not leak secrets (see \Cref{ex:lscct}).

\subsubsection{\gls*{ms}, Strict Cryptographic Constant Time}\label{sec:msscct-rel}

The combination of \gls*{ms} and \gls*{scct} is the intersection of these properties, \gls*{msscct}.
However, \gls*{ms} uses a different trace model than \gls*{scct}, so intersecting them would trivially yield the empty set. 
To remedy this issue, we introduce $\sim_{\ctsafe}: \preevent[_{\ctsafe}] \times \preevent[_{\mssafe}] $, a cross-language trace relation (whose key cases are presented below), that we use to intuitively unify the trace model in which the two properties are expressed:
\[
  \typerulenolabel{mscctrel:drop:tag}{}{\preevent[_{\mssafe}]\sim_{\ctsafe}\preevent[_{\mssafe}];\unlock{}}
  \typerulenolabel{mscctrel:drop:crash}{}{\lightning\sim_{\ctsafe}\lightning}
\]
\[
  \typerulenolabel{mscctrel:drop:branch}{}{\emptyevent\sim_{\ctsafe}\ev{Branch n};\securitytag{}}
  \typerulenolabel{mscctrel:drop:binop}{}{\emptyevent\sim_{\ctsafe}\ev{Binop n};\securitytag{}}
\]

Essentially, $\sim_{\ctsafe}$ ignores both the new $\ev{Branch\ n}$ and $\ev{Binop\ n}$ base-events as it relates security-insensitive actions ($\unlock$) to their equivalent counterparts.
Thus, \gls*{ms} traces trivially satisfy \gls*{scct}.
% 
The above relation is extended point-wise to traces, skipping the empty event $\emptyevent$ on either side, and it is now possible to define \gls*{msscct} using the universal image:

\begin{definition}[\gls*{ms} and \gls*{scct}]\label{def:trace:msscctdef}
  $
  \msscctsafe:=\mssafe\cap\sigma_{\sim_{\ctsafe}}\left(\scctsafe\right)
  $
\end{definition}

\subsection{Extending the Trace Model with Speculation}\label{subsec:msctss:tracemodel}

So far, the considered trace models do not let us express speculative execution attacks such as Spectre~\cite{kocher2019spectre}. 
For this, we extend the earlier trace model (see \Cref{subsec:scct:tracemodel}) so that the security tags ($\securitytag{}$) carry additional information about the kind of private data leakage, i.e., the type of speculative leak.
Moreover, we add base-events signalling the beginning of a speculative execution ($\ev{Spec}$), a barrier ($\ev{Barrier}$) that signals that any speculative execution may not go past it, as well as a rollback event ($\ev{Rlb}$), which signals that execution resumes to where speculation started.

\vspace{-1em}
{
\[
  \begin{array}{rrcl}
    (\text{Base-Events}) & \preevent[_{\mathghost}] &:=& \preevent[_{\ctsafe}] \\
    (\text{Spectre Variants}) & vX &:=& \operatorname{NONE} \mid \operatorname{PHT} \\
    (\text{Security Tags}) & \securitytag{} &:=& \lock_{vX} \mid \unlock\\ 
    (\text{Events}) & \event[_{\mathghost}] &:=& \preevent[_{\mathghost}];\securitytag{} \mid \emptyevent \mid \lightning \mid \ev{Spec} \mid \ev{Rlb} \mid \ev{Barrier}\\ 
  \end{array}
\]
}

Even though the considered Spectre variants are just SPECTRE-PHT~\cite{kocher2019spectre}, NONE just describes secret data as in \gls*{scct} (see \Cref{subsec:scct:tracemodel}), the trace model is general enough to allow for potential future extension with different variants~\cite{kocher2019spectre,maisuradze2018ret2spec,horn2019zero}.
%For sake of readability, this paper just uses the notation $\lock$ in place of $\lock_{\text{PHT}}$.

\subsubsection{Speculative Safety}

\gls*{ss}~\cite{patrignani2021exorcising}, similar to \gls*{scct}, is a sound overapproximation of a variant of noninterference.

\begin{definition}[\glsfirst*{ss}]\label{def:trace:ss}
  \noindent

  \begin{nscenter}
  $
    \sssafe:=\left\{\trace[_{\mathghost}] \left|\begin{array}{l}
      \trace[_{\mathghost}]=\hole{\cdot} \text{ or } \exists\trace[_{\mathghost}'].\\
      \left(\trace[_{\mathghost}]=\preevent[_{\mathghost}];\unlock\cdot\trace[_{\mathghost}'] \text{or }\trace[_{\mathghost}]=\preevent[_{\mathghost}];\lock_{\text{NONE}}\cdot\trace[_{\mathghost}']\right)\\
      \text{and }\ \trace[_{\mathghost}']\in\sssafe
                                 \end{array}\right.\right\}
  $ 
  \end{nscenter}
\end{definition}
The technical setup so far leads to the above definition, where only locks annotated with $\text{SPECTRE-PHT}$ are disallowed to occur on the trace.
That way, programs attaining \gls*{ss} do not necessarily attain \gls*{scct}.

\subsubsection{Speculation Memory Safety}\label{sec:spec-ms-rel}

As before, we need to relate the different trace models with each other, so that the memory safety property without speculation can be lifted to speculation. 
To this end, let $\sim_{\mathghost}: \preevent[_{\mathghost}] \times \preevent[_{\ctsafe}]$ be a cross-language trace relation whose key cases are below.
The intuition is that \gls*{ss} is trivially satisfied in \gls*{scct}, since speculation is inexpressible there, which amounts to dropping events $\ev{Spec}$, $\ev{Rlb}$, or $\ev{Barrier}$, as well as all base events tagged with $\lock_{\text{PHT}}$.

\[
  \typerulenolabel{specrel:drop:tag}{}{\preevent[_{\ctsafe}];\lock\sim_{\mathghost}\preevent[_{\mathghost}];\lock_{\text{NONE}}}
  \typerulenolabel{specrel:drop:spectag}{}{\emptyevent\sim_{\mathghost}\preevent[_{\mathghost}];\lock_{\text{PHT}}}
\]
\[
  \typerulenolabel{specrel:drop:pub}{}{\preevent[_{\ctsafe}];\unlock\sim_{\mathghost}\preevent[_{\mathghost}];\unlock}
  \typerulenolabel{specrel:drop:crash}{}{\lightning\sim_{\mathghost}\lightning}
\]
\[
  \typerulenolabel{specrel:drop:spec}{}{\emptyevent\sim_{\mathghost}\ev{Spec}}
  \typerulenolabel{specrel:drop:rlb}{}{\emptyevent\sim_{\mathghost}\ev{Rlb}}
  \typerulenolabel{specrel:drop:barrier}{}{\emptyevent\sim_{\mathghost}\ev{Barrier}}
\]

We conclude by defining the ultimate property of interest for secure compilers: \gls*{specms}.
\begin{definition}[\gls*{specms}]\label{def:trace:specmsdef}
  $
  \specmssafe := \msscctsafe\cap\sigma_{\sim_{\mathghost}}\left(\sssafe\right)
  $
\end{definition}

\section{Case Study: Language Formalisations\pages{2}}\label{sec:casestud:defs}

This section defines programming languages $\Ltms$, $\Ltrg$, $\Lms$, $\Lscct$, and $\Lspec$, all of which share common elements (presented in \Cref{subsec:cs:defs}).
$\Ltms$ is the only statically-typed language, and it exhibits the property that all well-typed programs are \gls*{tms} (\Cref{subsec:ltms}).
However, not all $\Ltms$ programs are \gls*{sms}.
That is, there are well-typed $\Ltms$ programs that perform an out-of-bounds access.
Language $\Ltrg$ is untyped and does not provide any guarantees with regards to \gls*{ms} (\Cref{subsec:lsms}).
$\Lms$ is exactly the same language as $\Ltrg$, but this paper still distinguishes the two for sake of readability (\Cref{subsec:lms}).
All three languages --- so $\Ltms$, $\Ltrg$, and $\Lms$ --- assume \gls*{scct} to hold, since this is -- in an ideal world -- what the programmer would expect, too: it is the job of the compiler to preserve and (potentially) enforce \gls*{scct} security~\cite{cauligi2019fact,nagarakatte2010cets,nagarakatte2009soft,akritidis2009baggy}.
% This paper considers \gls*{cct} as an architecture-dependent property\MPin{can we justify this}, so

Such consideration is also backed up by architecture providing a data (operand) independent timing mode, such as processors by Arm~\cite[p.~543]{arm-refman} and Intel~\cite[p.~80]{intel-refman}.
%In spirit of this, all languages besides $\Lscct$ are trivially \gls*{scct}. 
This kind of processor feature is modelled in language $\Lscct$ (\Cref{subsec:lscct}), where programs have access to a ``\gls*{cct}-mode'' and can change the leakage of emitted events according to the value of this mode (either $\obj{ON}$ or $\obj{OFF}$).

Finally, modern processors also employ speculative execution to achieve speedups---and unfortunately generate Spectre attacks~\cite{kocher2019spectre}---and this is the extension of $\Lspec$ (\Cref{subsec:lspec}).
 % extends $\Lscct$ to model speculative execution.
%
Thus, all previous languages trivially satisfy \gls*{ss}, since they do not support speculative execution at all.
% All languages share large parts of syntactic elements, which are presented in the next subsection (\Cref{subsec:cs:defs}).
% Subsequent subsections extend on this accordingly.

\subsection{Shared Language Definitions}\label{subsec:cs:defs}

All presented programming languages share a common fragment which is partially presented here and in full detail in the technical report. 
%\MP{put omitted stuff in appendix + ref there (and add textual explanation)}

% \vspace{-.5em}
{
  \renewcommand{\src}[1]{\mathsf{#1}}
 \small
\[
  \begin{array}{rrcl}
    \text{(Base-Events)} & \src{\preevent} &:=& \src{\ev{Alloc \loc\ n}} \mid \src{\ev{Dealloc \loc}} \mid \src{\ev{Get \loc\ n}} \mid \src{\ev{Set \loc\ n}} \\
    \text{(Control Tags)} & \src{\sandboxtag{}} &:=& \src{\ctx} \mid \src{\comp} \\
    \text{(Events)} & \src{\event} &:=& \src{\preevent;\sandboxtag{}} \mid \src{\emptyevent} \mid \src{\lightning} \\ 
    \text{(Values)} & \src{v} &:=& \src{n} \\
    \text{(Expressions)} & \src{\expr} &:=& \src{x} \mid \src{n} \mid \src{\lbinop{\expr[_1]}{\expr[_2]}} \mid \src{\lifz{\expr[_1]}{\expr[_2]}{\expr[_3]}} \\ 
                         &&&\mid \src{\llet{x}{\expr[_1]}{\expr[_2]}} \mid \src{\llet{x}{new\ \expr[_1]}{\expr[_2]}} \\
                         &&&\mid \src{\ldelete{x}} \mid \src{\lget{x}{\expr}} \mid \src{\lset{x}{\expr[_1]}{\expr[]}} \mid \src{stuck}\\
                         &&&\mid \src{\lcall{f}{\expr}} \mid \src{\lreturn{\expr}} \\
    \text{(Functions)} & \src{F} &:=& (\src{x};\src{\expr}) \\
    \text{(Libraries)} &&&\hspace{-3.8em} \src{\library_{\ctx}},\src{\library_{\comp}} : \src{Vars} \to \src{F} 
    % \hspace{2em}
    %  : \src{Vars} \to \src{F} 
     \\
    \text{(Heaps)} & \src{H} &:=& \src{\hole{\cdot}} \mid \src{n},\src{H}
    % \hspace{0.5cm}    
    \\
    \text{(Pointer Maps)} & \src{\Delta} && \text{ omitted for simplicity}
    \\
    \text{(Memory States)} & \src{\memstate} &:=& \left(\src{H^{\ctx}};\src{H^{\comp}};\src{\Delta}\right)\\
    \text{(Control States)} & \src{\cfstate} & & \text{ omitted for simplicity} 
    \\
    \text{(States)} &\src{\Omega} &:=& \left(\src{\cfstate};\src{\sandboxtag{}};\src{\memstate}\right)\\
  \end{array}
\]
}
{
  \renewcommand{\src}[1]{\mathsf{#1}}
The trace models of all languages are similar to those presented earlier (\Cref{sec:compprop}).
One technical detail is the addition of a control tag, indicating who is to blame for an emitted action: context ($\ctx$) or component ($\comp$).
This is a standard technique in secure compilation in order to rule out irrelevant context-level events. 
For example, a context immediately deallocating an allocated memory region twice trivially violates memory safety, but the main interest in secure compilation is to preserve component-level security, and thus component events.
Even though this tagging could be used for blame preservation~\cite{patrignani2023blame}, this is beyond the focus of this paper.
Another key difference is that memory accesses are now explicitly modelled as reads ($\src{\ev{Read\ \loc\ n}}$) and writes ($\src{\ev{Write\ \loc\ n}}$) instead of just uses.

>>>>

\PassOptionsToPackage{dvipsnames}{xcolor} % Added for new colors.
% Fixing: Too many math alphabets used in version normal
\newcommand\hmmax{0}
\newcommand\bmmax{0}
\documentclass[acmsmall]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

\input{sections/intro.tex}
\input{sections/background.tex}
\input{sections/compo.tex}
\input{sections/prop.tex}
\input{sections/langs.tex}
\input{sections/csc.tex}
